---
output: md_document
author: "Imke Mayer"
date: 2018-10-24
title: Complete list of selected references
weight: 2
url: "/bibliography/biblio_complete/"
---


```{r, load_refs, echo=FALSE, message=FALSE}
  require(dplyr)
  require(RefManageR)
  bibfile <- "../resources/rmisstastic_biblio.bib"
  bib   <- ReadBib(bibfile, check = "warn", .Encoding = "UTF-8")
  refs <- as.data.frame(bib)
  keys <- rownames(refs)
  refs["key"] <- keys
  
  # assign "categories" to the different types of publications
  refs <- refs %>%
      dplyr::mutate(
        pubtype = dplyr::case_when(bibtype == "Article" ~ "2",
                                 bibtype == "Article in Press" ~ "2",
                                 bibtype == "InProceedings" ~ "1",
                                 bibtype == "Proceedings" ~ "1",
                                 bibtype == "Conference" ~ "1",
                                 bibtype == "Conference Paper" ~ "1",
                                 bibtype == "MastersThesis" ~ "3",
                                 bibtype == "PhdThesis" ~ "3",
                                 bibtype == "Manual" ~ "4",
                                 bibtype == "TechReport" ~ "4",
                                 bibtype == "Book" ~ "5",
                                 bibtype == "InCollection" ~ "6",
                                 bibtype == "InBook" ~ "6",
                                 bibtype == "Misc" ~ "0",
                                 TRUE ~ "0"))
  
  ui <- "- "
  
  correct_topic <- function(x, tags_list){
    tags <- unlist(strsplit(x,"; "))
    return(length(intersect(tags_list,tags))>0)
  }
  
  select_refs <- function(refs, tags_list){
    bool_select <- apply(refs, FUN = function(x) correct_topic(unlist(x["topics"]), tags_list), MARGIN = 1)
    refs_selected <- refs %>%
      dplyr::mutate(select = bool_select) %>%
      filter(select)
    return(refs_selected)
  }
  
  print_ref_list_item <- function(ref, ui="- "){
      s <- paste(utils::capture.output(print(bib[key = ref["key"]], 
          .opts = list(check.entries = FALSE, 
                       style = "html", 
                       bib.style = "authortitle",
                       no.print.fields = c("issn","url","doi")))),collapse='')
      li_plus_datatype <- paste0('<li class="list-group-item" data-type="',
                            ref["bibtype"],
                            '">')
      s <- gsub('<p>', li_plus_datatype, s)
      s <- gsub("</p>", "", s)
      
      # change font colour for publication name
      s <- gsub("&ldquo;", '&ldquo;<font color="#428bca">',s)
      s <- gsub("&rdquo;", '</font>&rdquo;',s)
      
      if (ref["pubtype"] %in% c(0,3,4,5,6)) {
        s <- gsub("<em>", '<em><font color="#428bca">',s)
        s <- gsub("</em>", '</font></em>',s)
      }
    
      doi_url_button <- ""
      if (!is.na(ref["doi"])){
        doi_url_button <- paste0(doi_url_button,
                                 '<a href="https://doi.org/',
                                 ref["doi"],
                                 '" ',
                                 'target="_blank" role="button" class="btn btn-outline-secondary btn-sm" >DOI</a>')
      }
      if (!is.na(ref["url"])){
        doi_url_button <- paste0(doi_url_button,
                                 '<a href="',
                                 ref["url"],
                                 '" ',
                                 'target="_blank" role="button" class="btn btn-outline-secondary btn-sm" >URL</a>')
      }
      if (doi_url_button != ""){
        writeLines(paste0(s,"<div>",doi_url_button,"</div></li>"))
      } else {
        writeLines(s)
      }
  }
  
  pubtype_downsampled <- function(bibtype){
    pubtype = dplyr::case_when(bibtype == "Article" ~ "Article",
                                 bibtype == "Article in Press" ~ "Article",
                                 bibtype == "InProceedings" ~ "Paper",
                                 bibtype == "Proceedings" ~ "Paper",
                                 bibtype == "Conference" ~ "Paper",
                                 bibtype == "Conference Paper" ~ "Paper",
                                 bibtype == "MastersThesis" ~ "Misc",
                                 bibtype == "PhdThesis" ~ "Misc",
                                 bibtype == "Manual" ~ "Misc",
                                 bibtype == "TechReport" ~ "Misc",
                                 bibtype == "Book" ~ "Book",
                                 bibtype == "InCollection" ~ "Book",
                                 bibtype == "InBook" ~ "Book",
                                 bibtype == "Misc" ~ "Misc",
                                 TRUE ~ "Misc")
    return(pubtype)
  }
  
  print_ref_table_item <- function(ref){
      s <- paste(utils::capture.output(print(bib[key = ref["key"]], 
          .opts = list(check.entries = FALSE, 
                       style = "html", 
                       bib.style = "authortitle",
                       no.print.fields = c("doi","issn","url")))),collapse=' ')
      pubtype <- pubtype_downsampled(ref["bibtype"])                
      table_row <- paste0('<tr class="task-list-row" ',
                          'data-year="',ref["year"],'" ',
                          'data-pubtype="',pubtype,'"> ',
                          '<td>')
      s <- gsub('<p>', table_row, s)
      s <- gsub("</p>", "",s)
      
      # change font colour for publication name
      s <- gsub("&ldquo;", '<font color="#428bca">',s)
      s <- gsub("&rdquo;", '</font>',s)
      
      if (ref["pubtype"] %in% c(0,3,4,5,6)) {
        s <- gsub("<em>", '<em><font color="#428bca">',s)
        s <- gsub("</em>", '</font></em>',s)
      }
    
      doi_url_button <- ""
      if (!is.na(ref["doi"])){
        doi_url_button <- paste0(doi_url_button,
                                 '<a href="https://doi.org/',
                                 ref["doi"],
                                 '" ',
                                 'target="_blank" role="button" class="btn btn-outline-secondary btn-sm" >DOI</a>')
      }
      if (!is.na(ref["url"])){
        doi_url_button <- paste0(doi_url_button,
                                 '<a href="',
                                 ref["url"],
                                 '" ',
                                 'target="_blank" role="button" class="btn btn-outline-secondary btn-sm" >URL</a>')
      }
      if (doi_url_button != ""){
        writeLines(paste0(s,'<div>',doi_url_button,'</div>',
                          '</td> <td>',ref["year"],
                          '</td> <td>',pubtype,'</td>'))
      } else {
        writeLines(paste0(s,'</td> <td>',ref["year"],
                          '</td> <td>',pubtype,'</td>'))
      }
  }
```

*A commented and structured version of this bibliography can be found [here](../biblio_topics/).*

<div class="container">
<div class="row">
<table class="table">
<thead>
<tr class="filters">
<th>Publication type
<select id="pub-type-filter" class="form-control">
<option>All</option>
<option>Book</option>
<option>Article</option>
<option>Paper</option>
<option>Misc</option>
</select>
</th>
<th>Year
<select id="year-filter" class="form-control">
</select>
</th>
<th>Author
<input id="author-filter" class="form-control" type="text" placeholder="Search by author name..">
</th>
</tr>
</thead>
</table>

<div class="card bg-light filterable">
<table id="references-list-tbl" class="table table-hover">
<thead>
<tr>
<th>Citation</th>
<th>Year</th>
<th>Publication type</th>
</tr>
</thead>
<tbody>
```{r, print_all, results='asis', echo=FALSE}
invisible(apply(refs, FUN = function(x) print_ref_table_item(x), MARGIN=1))
```
</tbody>
</table>
</div>
</div>
</div>

<script>
{
var start = new Date().getFullYear();
var end = `r min(as.integer(refs$year))`;
var options = "<option>All</option>";
for(var year = start ; year >=end; year--){
  options += "<option>"+ year +"</option>";
}
document.getElementById("year-filter").innerHTML = options;
}
</script>

<script>
$(document).ready(function(){
  $("#author-filter").on("keyup", function() {
    var value = $(this).val().toLowerCase();
    $("#references-list-tbl tr").filter(function() {
      $(this).toggle($(this).text().toLowerCase().indexOf(value) > -1)
    });
  });
});
</script>

<script>
{
var
  filters = {
    pubtype: null,
    year: null
  };

function updateFilters() {
  $('.task-list-row').hide().filter(function() {
    var
      self = $(this),
      result = true; // not guilty until proven guilty
      Object.keys(filters).forEach(function (filter) {
      if (filters[filter] && (filters[filter] != 'None') && (filters[filter] != 'All')) {
        result = result && ((filters[filter] === self.data(filter)) || (filters[filter] === self.data(filter).toString()));
      }
    });

    return result;
  }).show();
}

function changeFilter(filterName) {
  filters[filterName] = this.value;
  updateFilters();
}

// Assigned User Dropdown Filter
$('#pub-type-filter').on('change', function() {
  changeFilter.call(this, 'pubtype');
});

// Year Dropdown Filter
$('#year-filter').on('change', function() {
  changeFilter.call(this, 'year');
});


/*
future use for a text input filter
$('#search').on('click', function() {
    $('.box').hide().filter(function() {
        return $(this).data('order-number') == $('#search-criteria').val().trim();
    }).show();
});*/
}
</script>

