---
output: md_document
author: "Steffen Moritz & Nathalie Vialaneix"
linktitle: Implementations
menu: navbar
title: Implementations
name: "Implementations"
url: "/rpkg/"
weight: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#library(imputeTS)
```

```{r PkGList, echo=FALSE, message=FALSE, warning=FALSE}
pkg_list <- c("imputeTS", "mice", "missForest", "missMDA", "naniar","simputation", "VIM", "Hmisc")
pkg_category <- c("Time-Series Imputation, Visualisations for Missing Data",
                  "Multiple Imputation",
                  "Single Imputation",
                  "Single and Multiple Imputation, Multivariate Data Analysis",
                  "Visualisations for Missing Data",
                  "Single Imputation, Meta-Package",
                  "Single Imputation, Visualisations for Missing Data",
                  "Single and Multiple Imputation, Data Processing")
```


We list some of the most popular R packages and Python libraries that allow to handle missing values.

<a href="#python_libraries">Click here to go directly to the list of Python libraries.</a>

#### R Packages

<p align="justify">Here are some introductions to popular missing data packages with small examples on how to use them. It gives more extensive information than the  <a href="https://CRAN.R-project.org/view=MissingData" target="_blank">CRAN Task View on Missing Data</a>, which is recommended to get a first overall overview about the CRAN missing data landscape.</p>

<p align="justify"><i>You can also contribute on your own to this page and provide a short introduction to a missing data package. Take a look at <a href="/packages/rpkg-listitem-template.md" target="_blank">this short description</a> on how to do this (<a href="/packages/rpkg-template.md" target="_blank">Template</a>). We are very happy about all contributions.</i></p>

<div id="users">
  <input class="search" placeholder="Search" />
  <button class="sort" data-sort="name">
    Sort by name
  </button>
  <button class="sort" data-sort="Category">
    Sort by Category
  </button>

<p>
  <ul class="list">
```{r loopRpkG, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
for (ind in seq_along(pkg_list)) {
    
  print_text <- paste0('<li style="margin: 0px;">',
                       '<h3 class="name">',
                       packageDescription(pkg_list[ind], fields = "Package"),
                       '</h3>',

                       '<h6 class="Category">Category: ',
                       pkg_category[ind],
                       '</h6>',
                       '<i class="Description">',
                       packageDescription(pkg_list[ind], fields = "Description"),
                       '</i><br>',
                       
                       '<img src="https://cranlogs.r-pkg.org/badges/',
                        pkg_list[ind],
                       '" alt="CRAN Downloads" />',
                       
                       '<img src="https://www.r-pkg.org/badges/last-release/',
                       pkg_list[ind],
                       '" alt="CRAN Release" /> <br>',
                       
                       '<a href="/packages/',
                       pkg_list[ind],
                       '"> more.. </a>',
                       
                       '</li></br>')
  cat(print_text)
}


```

<li style="margin: 0px;">
  <h3 class="name">imputomics</h3>
  <h6 class="Category">Category: Single and multiple imputation</h6> 
  <i class="Description">A robust wrapper package containing a range of methods for simulating and imputing missing values in different types of omics data such as genomics, transcriptomics, proteomics, and metabolomics. Provides tools for comparing and evaluating the performance of imputation methods and a web server.</i> <br> 
  <a href="/packages/imputomics"> more.. </a> 
</li> 

  </ul>
</p>
</div>

<script src="//cdnjs.cloudflare.com/ajax/libs/list.js/1.5.0/list.min.js"></script>
<script> var options = {
  valueNames: [ 'name', 'Category', 'Description' ]
};

var userList = new List('users', options);
</script>


#### Python modules {#python_libraries}


<p align="justify">Here are some links to modules or methods in Python to handle missing values.


* [sklearn.impute](https://scikit-learn.org/stable/modules/classes.html#module-sklearn.impute): module from sklearn for missing value imputation (simple imputation, conditional iterative imputer, k-Nearest Neighbors imputer).

* [pandas](https://pandas.pydata.org/pandas-docs/stable/user_guide/missing_data.html): available methods in pandas to handle dataframes with missing values (fill the missing values by a constant, remove missing values).

* [statsmodels.imputation](https://www.statsmodels.org/stable/imputation.html): module from statsmodels to handle missing values (multiple imputation, Bayesian imputation using a Gaussian model).

</p>

